GrapeCPNet: A Self-supervised Point Cloud Completion Network for 3D Phenotyping of Grape Bunches

### 2025-05-02 response2reviewer.tex

    When resubmitting your manuscript, please carefully consider all issues mentioned in the reviewers' comments, outline every change made point by point, and provide suitable rebuttals for any comments not addressed.
\end{reviewercomment}

\response{
    We sincerely appreciate the positive feedback from the editor and reviewers regarding our manuscript. 
    We are grateful for the opportunity to revise our work based on the valuable comments received.

    In this revision, we have:

    1. Carefully addressed all reviewer comments point-by-point;\\
    2. ...
    % 2. Provided detailed responses explaining each modification;\\
    % 3. Made all necessary revisions to improve manuscript clarity and technical accuracy;\\
    % 4. Included tracked changes in the revised manuscript for transparent review.

@@

    ...
}

\todoblock{
    Summary of added experiment.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{reviewercomment}
    Regarding the instance segmentation, I understand that this is not the main contribution of the paper. Still, it would be nice to test different approaches here to evaluate the robustness of the completion network to varying levels of wrong predictions. Along a similar line, I would have liked to see the completion network working on ground truth segmentation labels as this will disentangle errors coming from the instance segmentation net and the completion net.
\end{reviewercomment}

\response{
    We sincerely appreciate the reviewer's valuable suggestions regarding the instance segmentation evaluation. 
    In response, we have conducted additional comparative experiments by applying our completion network to both manually annotated ground truth labels and SoftGroup segmentation results, with quantitative comparisons shown in the new results figure and table below.

    A t-test analysis revealed no statistically significant differences between the two approaches (p > xxx), demonstrating that our high-accuracy segmentation network (achieving mAP@[.5:.95] of 0.995 represents an AP50 of 0.999 for our results) provides sufficiently reliable inputs for the completion network.

    We maintain our design of having specialized networks focus on their respective tasks: the segmentation network achieves reliable segmentation accuracy, while the completion network specializes in partial point cloud completion. 
    Rather than increasing the completion network's robustness to handle poor segmentation results.
    It would require generating artificial training data with various error patterns that might negatively impact training.
    We believe it is more principled to improve the segmentation performance directly. 
    This modular approach ensures clear responsibility separation between network components.
}

\todoblock{
    Execute the following experiment on manual annotation bunches:

    1. Apply completion networks directly on ground truth (separated 3D reconstruction)

    2. Run SoftGroup segmentation on ground truth, and then apply completion networks on these results

    3. Compare two results to draw comparison tables and runing t-test to show no significant differences.
}

@@

\manuscript{
    ...
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{reviewercomment}
    Regarding the completion net, I would have liked to see ablation studies here as I believe this is the main  contribution of this work. For example, how does the number of cuts  influence the results of the completion metrics? How does the number of  output points in the centroid-based contour influence the snowflake net?
\end{reviewercomment}

\response{
    We thank the reviewer for these insightful suggestions regarding ablation studies for our completion network. 
    For current manuscript, we already have part ablation studies. 
    The GrapeCPNet is combined the modification of PoinTr and SnowFlakeNet.
    As shown in the original Table 3, we compared the performance of:
    (1) Standalone PoinTr; (2) Standalone SnowFlakeNet; (3) Our modified GrapeCPNet (combined architecture).
    The results demonstrated the superiority of our integrated approach.

    We have also expanded our ablation studies to provide more comprehensive analysis as you suggested:

    1. \textbf{Impact of Cut Numbers}: We conducted new experiments evaluating how different occlusion levels (number of cuts) affect completion performance:

    ...

    The results showed that mixed occlusion training significantly improves generalization

    2. \textbf{Point Number Analysis}: We investigated how output point count affects SnowFlakeNet performance:

    ...

    The result showed higher point counts yield better completion quality.
}

\todoblock{
    To add the experiement:

    1. split training data to 1 cuts, 2cuts, 3cuts, 4 cuts and 5 cuts, train 5 individual networks and validate on val dataset and compare with 1-5 full training data? (expected -> less cuts in training data, worse performance on server overlapped.)

    2. Decrease the feature M (current=2048) of networks, 2048, 1024, 512, 256, to see if any affects of completion results.
}

@@

\manuscript{
    ...
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{reviewercomment}
    Finally, is there a reason why the approach of Blok et al. 2025 is not being used as a  baseline? I would have liked to see a non-learning-based baseline as  well, like Marangoz et al. 2022
\end{reviewercomment}

\response{
    We appreciate the reviewer's suggestion regarding additional baselines.

    The work of \citet{marangoz_fruit_2022} generates cube-like point clouds for peppers, which may not be suitable for our sphere-like grape berries.
    Additionally, re-implementation their algorithm is time-consuming and unlikely to benefit our current grape project.
    Our ellipsoid surface fitting is a similar non-learning-based approach, serving for berry traits analysis in this study. 
    Therefore, it may not be appropriate as a baseline comparison.

    The comparison with \citet{blok_highthroughput_2025} was not included initially because these were independent parallel research projects during our manuscript preparation phase. 
    Following editorial recommendations during the submission process, we subsequently added discussions of comparable works in the revised manuscript, including it into the discussion part after its acceptance.

    In this revision, we have now included CoRe++ from \citet{blok_highthroughput_2025} performance on our grape dataset in Table 3. 
    However, we observe that the network architectures target substantially different objectives.
    Our grape dataset involves partial completion from the same sensor type, while the potato network focuses on fusing different sensors 
    (with incomplete point clouds from RGB-D and complete ones from SfM, exhibiting significant shape and texture differences). 
    Our approach emphasizes better completion performance while the \citet{blok_highthroughput_2025} more on processing speed and real-time processing, and consequently, CoRe++ does not demonstrate significant advantages in our grape dataset scenario.
}

\todoblock{
    Add CoRe++ results to Table 3 (collaboration with pieter)
}

@@

\manuscript{
    In order to train and validate SoftGroup, each berry and stem in the bunch point cloud obtained in Section~\ref{sec:212} was \added[id=R2]{individually} labeled \replaced[id=R2]{(instance-level segmentation) as shown in}{at the instance-level} Fig.~\ref{fig:raw11}.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{reviewercomment}
    Which platform, software, or tool was used for labeling?
\end{reviewercomment}

\response{
    We thank the reviewer for raising this important methodological detail. 
    The 3D point clouds were labeled using CloudCompare (v2.12.4), an open-source software for 3D point cloud and mesh processing.
}

\todoblock{
    Confirm which tool for labeling, or SoftGroup can automatically annotation?
}

### 2025-05-08 response2reviewer.tex

    1. Carefully addressed all reviewer comments point-by-point;\\
    2. ...
    % 2. Provided detailed responses explaining each modification;\\
    % 3. Made all necessary revisions to improve manuscript clarity and technical accuracy;\\
    % 4. Included tracked changes in the revised manuscript for transparent review.

    We believe these revisions have significantly strengthened the paper, and we appreciate the editor's and reviewers' time and constructive suggestions. 
    Please find our point-by-point responses to each reviewer comment in the following sections.
}

%============
% Reviewer 1
%============
\reviewer
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{reviewercomment}
    Overall, the paper is well structured with a clear goal, and the authors provide insight to justify their design choices. 
    However, I believe the paper will benefit from  more experiments, particularly on the instance segmentation and completion parts.
\end{reviewercomment}

\response{
    We sincerely appreciate the reviewer's valuable feedback and constructive suggestions for improving our paper. 
    Regarding the request for additional experiments on instance segmentation and completion,
    We have enhanced the manuscript to better explain these design choices in the Methods, Results and  Discussion sections.
    Please find our detailed responses to each specific comment below.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{reviewercomment}
    Regarding the instance segmentation, I understand that this is not the main contribution of the paper. Still, it would be nice to test different approaches here to evaluate the robustness of the completion network to varying levels of wrong predictions. Along a similar line, I would have liked to see the completion network working on ground truth segmentation labels as this will disentangle errors coming from the instance segmentation net and the completion net.
\end{reviewercomment}

\response{
    We sincerely appreciate the reviewer's insightful comments regarding the evaluation of instance segmentation and its impact on the completion network. 
    We agree that analyzing the robustness of the completion network to segmentation errors is an interesting direction, and we already have conducted a similar preliminary experiments during the early stage of our studies.

    In our preliminary tests on a small subset of data, we visually compared the outputs of completion network on using ground truth segmentation versus SoftGroup segmentation outputs. 
    The differences were negligible to the human eye thus not mentioned in the manuscript.
    This is because our high-quality grape point clouds (captured in a controlled indoor environment with high-overlap scanning) exhibit clear boundaries and minimal noise. 
    Also, since our segmentation network (trained on manually annotated data using SoftGroup) achieves near-perfect performance (mAP@[.5:.95] of 0.995, representing AP50 of 0.999), making its outputs virtually almost indistinguishable from ground truth for our application.
    Moreover, we note that manual annotation of 3D data on 2D displays inevitably contains some human error, meaning even ``ground truth'' segmentations are not perfectly accurate.

    Given these observations, we believe our current design - where each specialized network focuses on its respective task - represents the most efficient approach. 
    The segmentation network reliably handles instance separation, while the completion network specializes in partial point cloud completion.
    While artificially generating poor segmentation outputs to test completion robustness could be interesting, 
    it would require additional design to generating extra training data to cover various error patterns that might negatively impact network training.

    Instead, we believe it's more principled to:
    1) Improve data collection quality, as poor-quality inputs lead to poor-quality outputs; and
    2) Enhance segmentation accuracy to ensure high-quality inputs for the completion network.
}

\manuscript{
    and the result of obtaining the phenotypic traits.

    \textbf{3.1 Berry Instance Segmentation Results}

    \added[id=R1]{
    Our controlled indoor environment (Fig.~\ref{fig:raw9}) with high-overlap scanning enabled high-quality grape point cloud reconstruction, featuring clear boundaries and minimal noise.
    Some reconstructed results are shown in Figs.~\ref{fig:raw10}, \ref{fig:raw2}, \ref{fig:raw11}, and \ref{fig:raw13}.
    These high-quality input data reduced the challenges in annotating instance berry segmentation training data.
    They also simplified the segmentation model training by minimizing concerns about noise and unclear boundaries.
}

    \added[id=R1]{The segmentation task divides the grape bunch into two parts: berries and stems.} 
    The mAP@[.5:.95] values of berry and stem class are 0.995 and 0.941, respectively, which showed a high accuracy of the berry segmentation from the bunch.
}

\manuscript{
    The primary time-consuming work of our approach was the annotation of 3D data for instance segmentation.
    \added[id=R1]{
        We acknowledge that manual annotation of 3D data on 2D displays may introduce some human errors, meaning even the ground truth segmentation annotations are not perfectly accurate.
        However, due to the high-quality 3D reconstruction with clear berry boundaries and minimal noise effects,
        we could train a highly accurate SoftGroup segmentation network, achieving an mAP@[.5:.95] of 0.995 and AP50 of 0.999.
    }
    Compared to similar studies on grape berries, our proposed method demonstrated superior performance, achieving the best results among the reviewed methods (Subsection~\ref{sec:insegresults}).
    \added[id=R1]{Instead of developing complex algorithms to handle low-quality inputs, improving input quality simplifies the problem and produces better outputs.}

    For berry completion using deep learning, the most challenging and time-consuming task is
}

\manuscript{
    \added[id=R2]{
        and assess whether models trained on table grapes can generalize well to smaller wine grape berries without additional training.
    }

    \added[id=R1]{
        The design philosophy focuses on ensuring input quality and assigning specialized networks to their respective tasks, which provides the most efficient approach. 
        Rather than developing complex algorithms to process low-quality inputs with unclear berry boundaries and noisy 3D models, improving input quality simplifies the problem and enhances the results. 
        Similarly, assigning each specialized network to its specific task is essential. 
        In this study, the segmentation network reliably performs instance separation, while the completion network specializes in partial point cloud completion. 
        Although the automatic training data generation strategy can be adjusted to produce poor segmentation outputs—for example, by adding noise or partially cutting neighboring berries—it remains difficult to cover all potential error patterns to improve robustness. 
        Additionally, training with too many low-quality samples may negatively affect network performance. 
        However, this design philosophy also faces challenges in practical applications.
    }
    \replaced[id=R1]{For example, this study}{Another limitation of this study is that it} focused on a single berry bunch in a controlled environment.
}

\manuscript{
    the branching structure could potentially be captured during early stages of thinning, as suggested by \citep{du_instance_2023}.
    \added[id=R1]{Lastly, with the development of lidar scanning techniques, some companies also provide low-cost products to improve 3D scanning results in the field, such as XGrids (XGRIDS LIMITED, \url{https://www.xgrids.com}).}
    Such data could enable studies on growth prediction for optimizing harvesting times and assessments of branching structure deviations within crowded berry bunches.
}

### 2025-05-08 main.tex
R1 point 2

The mAP@[.5:.95] values of berry and stem class are 0.995 and 0.941, respectively, which showed a high accuracy of the berry segmentation from the bunch.
Since mAP@[.5:.95] is a strict evaluation metric, it represents the mean values of Average Precision (AP) scores calculated from AP50 to AP95 at 5\% IoU intervals.
Compared to other studies, \citet{ni_threedimensional_2021} reported that their proposed segmentation network achieved a mAP@[.5:.95] of 0.877 on their testing dataset, while \citet{du_instance_2023} achieved a mAP[.5:.95] of 0.657 and an AP50 of 0.952.
In our study, the AP50 value exceeded 0.999, with both recall and precision values at this level also over 0.999.
When compared to traidtional berry semgnetation algorithms, \citet{rose_automated_2016} reported a mean recall of 0.776 and precision of 0.978;
The results of four fresh grapes for bunch point cloud segmentation are visualized in Fig.~\ref{fig:raw13}.
According that figure, we can also see an almost perfect instance berry segmentation results.

@@

% revisit gap/aim/method
For individual berry segmentation, early studies relied on the circular characteristics of the visible portions of berries for segmentation.
Building on the work of \citet{schnabel_efficient_2007}, who introduced point cloud shape detection using RANSAC, researchers were able to cluster partial berries based on their rounded shapes \citep{scholer_automated_2015, rist_highprecision_2018, schneider_predicting_2020}.
However, this feature-based approach performs well only in sparse grape arrangements and proves ineffective in more complex, dense conditions.
The reliance on manually designed features for algorithm development has now reached its limitations.
With advancements in machine learning, the focus has shifted to a data-driven era, where algorithms can learn complex feature patterns automatically.
For instance, \citet{rose_automated_2016} applied machine learning to distinguish features between grape berry clusters and background leaves automatically.
Similarly, \citet{ni_threedimensional_2021} implemented a two-step approach by first using mature 2D instance segmentation techniques on 2D images and then projecting the resulting segmentation masks onto reconstructed 3D models to complete the challenging task of 3D instance segmentation.
In another approach, \citet{luo_3d_2023} Mask R-CNN for 2D image segmentation and projected the results onto RGB-D point clouds.
The normal directions derived from the point clouds were then packed as an additional feature in the PointNet++ network for point cloud segmentation.
In this study, we applied the state-of-the-art point cloud segmentation network, SoftGroup \citep{vu_softgroup_2022} directly to the problem.
The primary time-consuming work of our approach was the annotation of 3D data for instance segmentation.
Compared to similar studies on grape berries, our proposed method demonstrated superior performance, achieving the best results among the reviewed methods (Subsection~\ref{sec:insegresults}).

@@

\replaced[id=R2]{Additionally, this study only}{Another limitation of this study is that it} focused on a single berry bunch in a controlled environment.
The noise in the data is also related to the effectiveness of the method proposed in this paper.
For practical applications in vineyards, collecting the same point cloud quality as that achieved in controlled indoor environments can be challenging, and multi-view data collection may not be as efficient.
How to effectively implement the method in practical application scenarios is another issue to be considered in future research.
To increasing data collection efficiency, devices such as RGB-D cameras \citep{blok_highthroughput_2025}, fisheye cameras \citep{tamura_unsupervised_2025}, or even 360 cameras can be used.
Moreover, as in previous studies \citep{rose_automated_2016, luo_3d_2023}, investigating methods to segment individual grape bunches directly from vineyard scans using deep learning techniques should be an area of future exploration.
Additionally, \citet[Fig.~9]{scholer_automated_2015} successfully demonstrated the construction of virtual 3D models of grape bunches, including the hidden branching architecture. 
By collecting time-series growth data, the branching structure could potentially be captured during early stages of thinning, as suggested by \citep{du_instance_2023}.
Such data could enable studies on growth prediction for optimizing harvesting times and assessments of branching structure deviations within crowded berry bunches.
Finally, applying the latest procedural computer graphics (CG) modeling techniques or functional-structural plant models represents another promising approach to advancing practical applications in this field.

>>>>

\added[id=R1]{
    Our controlled indoor environment (Fig.~\ref{fig:raw9}) with high-overlap scanning enabled high-quality grape point cloud reconstruction, featuring clear boundaries and minimal noise.
    Some reconstructed results are shown in Figs.~\ref{fig:raw10}, \ref{fig:raw2}, \ref{fig:raw11}, and \ref{fig:raw13}.
    These high-quality input data reduced the challenges in annotating instance berry segmentation training data.
    They also simplified the segmentation model training by minimizing concerns about noise and unclear boundaries.
}

\added[id=R1]{The segmentation task divides the grape bunch into two parts: berries and stems.} 
The mAP@[.5:.95] values of berry and stem class are 0.995 and 0.941, respectively, which showed a high accuracy of the berry segmentation from the bunch.
Since mAP@[.5:.95] is a strict evaluation metric, it represents the mean values of Average Precision (AP) scores calculated from AP50 to AP95 at 5\% IoU intervals.
Compared to other studies, \citet{ni_threedimensional_2021} reported that their proposed segmentation network achieved a mAP@[.5:.95] of 0.877 on their testing dataset, while \citet{du_instance_2023} achieved a mAP[.5:.95] of 0.657 and an AP50 of 0.952.
In our study, the AP50 value exceeded 0.999, with both recall and precision values at this level also over 0.999.
When compared to traditional berry segmentation algorithms, \citet{rose_automated_2016} reported a mean recall of 0.776 and precision of 0.978;
The results of four fresh grapes for bunch point cloud segmentation are visualized in Fig.~\ref{fig:raw13}.
According that figure, we can also see an almost perfect instance berry segmentation results.

@@

% revisit gap/aim/method
For individual berry segmentation, early studies relied on the circular characteristics of the visible portions of berries for segmentation.
Building on the work of \citet{schnabel_efficient_2007}, who introduced point cloud shape detection using RANSAC, researchers were able to cluster partial berries based on their rounded shapes \citep{scholer_automated_2015, rist_highprecision_2018, schneider_predicting_2020}.
However, this feature-based approach performs well only in sparse grape arrangements and proves ineffective in more complex, dense conditions.
The reliance on manually designed features for algorithm development has now reached its limitations.
With advancements in machine learning, the focus has shifted to a data-driven era, where algorithms can learn complex feature patterns automatically.
For instance, \citet{rose_automated_2016} applied machine learning to distinguish features between grape berry clusters and background leaves automatically.
Similarly, \citet{ni_threedimensional_2021} implemented a two-step approach by first using mature 2D instance segmentation techniques on 2D images and then projecting the resulting segmentation masks onto reconstructed 3D models to complete the challenging task of 3D instance segmentation.
In another approach, \citet{luo_3d_2023} Mask R-CNN for 2D image segmentation and projected the results onto RGB-D point clouds.
The normal directions derived from the point clouds were then packed as an additional feature in the PointNet++ network for point cloud segmentation.
In this study, we applied the state-of-the-art point cloud segmentation network, SoftGroup \citep{vu_softgroup_2022} directly to the problem.
The primary time-consuming work of our approach was the annotation of 3D data for instance segmentation.
\added[id=R1]{
    We acknowledge that manual annotation of 3D data on 2D displays may introduce some human errors, meaning even the ground truth segmentation annotations are not perfectly accurate.
    However, due to the high-quality 3D reconstruction with clear berry boundaries and minimal noise effects,
    we could train a highly accurate SoftGroup segmentation network, achieving an mAP@[.5:.95] of 0.995 and AP50 of 0.999.
}
Compared to similar studies on grape berries, our proposed method demonstrated superior performance, achieving the best results among the reviewed methods (Subsection~\ref{sec:insegresults}).

@@

\added[id=R1]{
    The design philosophy focuses on ensuring input quality and assigning specialized networks to their respective tasks, which provides the most efficient approach. 
    Rather than developing complex algorithms to process low-quality inputs with unclear berry boundaries and noisy 3D models, improving input quality simplifies the problem and enhances the results. 
    Similarly, assigning each specialized network to its specific task is essential. 
    In this study, the segmentation network reliably performs instance separation, while the completion network specializes in partial point cloud completion. 
    Although the automatic training data generation strategy can be adjusted to produce poor segmentation outputs—for example, by adding noise or partially cutting neighboring berries—it remains difficult to cover all potential error patterns to improve robustness. 
    Additionally, training with too many low-quality samples may negatively affect network performance. 
    However, this design philosophy also faces challenges in practical applications.
}
\replaced[id=R1]{For example, this study}{Another limitation of this study is that it} focused on a single berry bunch in a controlled environment.
The noise in the data is also related to the effectiveness of the method proposed in this paper.
For practical applications in vineyards, collecting the same point cloud quality as that achieved in controlled indoor environments can be challenging, and multi-view data collection may not be as efficient.
How to effectively implement the method in practical application scenarios is another issue to be considered in future research.
To increasing data collection efficiency, devices such as RGB-D cameras \citep{blok_highthroughput_2025}, fisheye cameras \citep{tamura_unsupervised_2025}, or even 360 cameras can be used.
Moreover, as in previous studies \citep{rose_automated_2016, luo_3d_2023}, investigating methods to segment individual grape bunches directly from vineyard scans using deep learning techniques should be an area of future exploration.
Additionally, \citet[Fig.~9]{scholer_automated_2015} successfully demonstrated the construction of virtual 3D models of grape bunches, including the hidden branching architecture. 
By collecting time-series growth data, the branching structure could potentially be captured during early stages of thinning, as suggested by \citep{du_instance_2023}.
\added[id=R1]{Lastly, with the development of lidar scanning techniques, some companies also provide low-cost products to improve 3D scanning results in the field, such as XGrids (XGRIDS LIMITED, \url{https://www.xgrids.com}).}
Such data could enable studies on growth prediction for optimizing harvesting times and assessments of branching structure deviations within crowded berry bunches.
Finally, applying the latest procedural computer graphics (CG) modeling techniques or functional-structural plant models represents another promising approach to advancing practical applications in this field.
